---
title: "01-Intro"
title: "01-Intro"
author: 'Matt Niederhuber'
date: '2/19/25'
date: '2/19/25'
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: show
---
<style>
body {
text-align: justify}
</style>

The goal of this section is to make sure everyone is roughly on the same page before diving into analysis.

We'll try to quickly cover some R basics that we'll use throughout the course. 
But our focus is not going to be on all of the details of R so we'll skip over a lot of the details.

# R Notebooks 

This is an R notebook. It is a mix of code and text. 
It allows you to create clear documentation of your code in markdown and
interactively run code in R. (or Python or Bash or etc...) 

Code is written in "chuncks". We can make a new chunk by clicking the "+" button in the toolbar.
Or with the keyboard shortcut `Ctrl + Alt + I`.
Or simply by typing 3 backticks "`" and then "{r}".

```{r}
# This is an R code chunk
# we can comment with hashtags
print("Hello World")
```

You end chunks with another 3 backticks.
To run a chunck click the green play button in the top right of the chunk.
Or use the keyboard shortcut `Ctrl + Enter`.

# Assigning Variables

There are 2 operators for assigning variables in R. 
'<-' and '='. 
They are functionally equivalent.

x <- 10 \
y = 10 \
are the same.

```{r}
x <- 10
y = 10
print(x)
print(y)
```

In our sections you'll see that I try to use "<-" outside of functions,
and "=" inside of functions. 

# Data Structures
There are 5 core data structures in R.

- **Atomic Vectors** are 1d vectors of the same data type.
- **Lists** are also 1d vectors but can have a mix of data types.
- **Matrices** are 2d (columns and rows) structures where all values are the same data type.
- **Dataframes** are 2d structures that can have a mix of data types. 
- **Arrays** are higher order n-dimensional structures of the same data type.

These all differ in their dimensionaly (1d, 2d, Nd) and the data that's stored in them (all the same type or a mix of types).\

Briefly, datatypes in R include decimal numbers (double), integers, letters/words (character), factors, and True/False (logical). \


### Vectors

We use the `c()` function to create vectors.
```{r}
x <- c(1, 2, 3, 4, 5) # a vector of numbers
typeof(x)
y <- c("a", "b", "c", "d", "e") # a vector of characters
typeof(y)
x <- c(1, 2, 3, 4, 5) # a vector of numbers
typeof(x)
y <- c("a", "b", "c", "d", "e") # a vector of characters
typeof(y)
```

And we can access elements of vectors with square brackets.
And we can access elements of vectors with square brackets.
```{r}
x[1] # the first element of x
y[3] # the third element of y

x[1:3] # the first 3 elements of x

x[x==3] # all elements of x that are equal to 3
```

There are a set of functions available to change the type of a vector. 
x[1] # the first element of x
y[3] # the third element of y

x[1:3] # the first 3 elements of x

x[x==3] # all elements of x that are equal to 3
```

There are a set of functions available to change the type of a vector. 
```{r}
as.character(x) # convert x to a character vector
```

If we try to convert a vector of characters to numeric all values become NA.
```{r}
as.numeric(y) # convert y to a numeric vector
```

There's also `as.logical()`, `as.factor()`, and `as.integer()`.

### Factors

Factors are a special type of vector that is useful for categorical data.
They are the same as vectors except they have "levels".
A factors' "levels" are all the unique categories/instances in the vector.
### Factors

Factors are a special type of vector that is useful for categorical data.
They are the same as vectors except they have "levels".
A factors' "levels" are all the unique categories/instances in the vector.

```{r}
factor(c("A","A","B","C","C","C")) # a factor with 3 levels: A, B, C
factor(c("A","A","B","C","C","C")) # a factor with 3 levels: A, B, C
```

### Lists
### Lists
```{r}
# we use list() to make lists
z <- list(1, "a", 2, "b") # lists can have a mix of data types
typeof(z)
# we use list() to make lists
z <- list(1, "a", 2, "b") # lists can have a mix of data types
typeof(z)
```

### Dataframes
Dataframes are the most common data structure in R. 
They are basically a list of equal length vectors. 
### Dataframes
Dataframes are the most common data structure in R. 
They are basically a list of equal length vectors. 

You can create a dataframe with the `data.frame()` function.
You can create a dataframe with the `data.frame()` function.
```{r}
# a dataframe with 3 columns, and 4 rows
df <- data.frame(names = c('Matt','Sally','Emily','Tracey'), 
                 age = c(36,35,2,5), 
                 favorite_food = c('tacos','fries','apples','candy'))
# a dataframe with 3 columns, and 4 rows
df <- data.frame(names = c('Matt','Sally','Emily','Tracey'), 
                 age = c(36,35,2,5), 
                 favorite_food = c('tacos','fries','apples','candy'))
df
```

Because a dataframe is a list of vectors, you can access columns with the `$` operator.
Because a dataframe is a list of vectors, you can access columns with the `$` operator.
```{r}
df$a
```

You can subset dataframes with square brackets.
df$a
```

You can subset dataframes with square brackets.
```{r}
df[1,] # the first row of df
df[,1] # the first column of df

df['names'] # the column named 'name'

df[df$names == 'Matt',] # all rows where column a is 'Matt'
```

We just used the `==` operator to compare values.
Other operators in R include `!=`, `>`, `<`, `>=`, `<=`.

There is also the `%in%` operator to check if a value is in a vector.

# Functions

We'll primarily be using other packages and their functions in this course.
We've already run some functions like `print()`, `c()`, and `data.frame()`.

The basic syntax for a function is `function_name(arguments)`, whith multiple arguments separate by ","

# Packages
A package in R is basically a collection of functions. There are a two main public repositories for R packages that you'll commonly use: 

- "The Comprehensive R Archive Network" CRAN https://cran.r-project.org/ 
- "Bioconductor" https://www.bioconductor.org/ -- specific for bioinformatics 

We can install new packages with the function `install.packages()`. Downloaded packages are stored in a system "library", by default R makes a library in your $HOME directory. 

Something like:
`/nas/longleaf/home/{ID}/R/x68_64-pc-linux-gnu-library/4.4`. 
`/nas/longleaf/home/{ID}/R/x68_64-pc-linux-gnu-library/4.4`. 

```{r, eval=FALSE}
install.packages('PACKAGENAME')
# and
remove.packages('PACKAGENAME')
# and
remove.packages('PACKAGENAME')
```

There are a couple of ways to load packages and use their functions. 
The most common way is to use the `library()` function.
There are a couple of ways to load packages and use their functions. 
The most common way is to use the `library()` function.
```{r}
library(dplyr)
library(dplyr)
```

The other option is to use the `::` operator.
The other option is to use the `::` operator.
```{r}
dplyr::select(df, age)
```

# Pipes

Pipes are a way to chain functions together. 
The output of one function is "piped" as the input to the next function.

There are 2 main pipe operators in R. 
`|>` from base R (as of R > 4.1.0)
`%>%` from the `magrittr` package. 

They are essentially the same, with some small differences.
I use `%>%` out of habit. 
dplyr::select(df, age)
```

# Pipes

Pipes are a way to chain functions together. 
The output of one function is "piped" as the input to the next function.

There are 2 main pipe operators in R. 
`|>` from base R (as of R > 4.1.0)
`%>%` from the `magrittr` package. 

They are essentially the same, with some small differences.
I use `%>%` out of habit. 

```{r}
library(magrittr)

df %>% 
  dplyr::mutate(tacos = ifelse(favorite_food == 'tacos', T, F)) 
```

# Resources

I have a more indepth notebook on R basics here [Intro to OnDemand and R](01-Intro/01-Intro.html)

df %>% 
  dplyr::mutate(tacos = ifelse(favorite_food == 'tacos', T, F)) 
```

# Resources

I have a more indepth notebook on R basics here [Intro to OnDemand and R](01-Intro/01-Intro.html)

[RStudio Education](https://education.rstudio.com/)

[R for Data Science](https://r4ds.had.co.nz/)

[Tidyverse](https://www.tidyverse.org/)

[R Graph Gallery](https://r-graph-gallery.com/)

[ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)

[Posit Cheatsheets](https://posit.co/resources/cheatsheets/)